Deep Learning

### 2015-01-31 deepreview.tex
many small changes and a big reorganization of sections

So far, most applications of deep learning use feed-forward neural
networks, illustrated in Figure~\ref{fig:backprop-box}, which learn
to map a fixed-size input (e.g., an image) to a fixed-size output
(e.g. a probability for each one of several candidate object categories). 
They do so by successive transformations corresponding to different
layers of artificial neurons (called {\em hidden units}),
which can be interpreted as learned derived features of the raw input. However, 
for tasks that involve sequential inputs, such as speech
and language, it is often better to use recurrent neural
networks, illustrated in Figure~\ref{fig:rnn} and discussed in Section~\ref{sec:rnn}. 
Recurrent neural networks process an input sequence one element at a time,
maintaining in their hidden units a ``state vector'' that implicitly contains information
about the history all the past elements of the sequence.
Recurrent networks are very good at predicting the next
character in text~\citep{Sutskever-et-al-ICML2011} or the next
word in a sequence~\citep{Mikolov-et-al-NIPS2013} but they can also be
used for more complex tasks.  After reading an English sentence one
word at a time, for example, an English ``encoder'' network can be
trained so that the final state vector of its hidden units is a good
representation of the thought expressed by the sentence.  This thought
vector can then be used as the initial hidden state of (or as extra
input to)  a jointly trained French ``decoder'' network which outputs a
probability distribution for the first word of the French
translation. If a particular first word is chosen 
from this distribution and provided as input to the decoder network it will
then output a probability distribution for the second word of the
translation and so on until a full stop is
chosen~\citep{Bahdanau-et-al-arxiv2014,Sutskever-et-al-NIPS2014}.
Overall, this process generates sequences of French words according to
a probability distribution that depends on the English sentence.

@@

\begin{figure}[htp]
\fbox{
\begin{minipage}{\textwidth}
\centerline{
\fbox{\includegraphics[width=0.69\linewidth]{netvis-simple-2S.png}}
\fbox{\includegraphics[width=0.28\linewidth]{chain-rule.pdf}}}
\centerline{
\hspace*{.5mm}\fbox{\includegraphics[width=0.49\linewidth]{bpfig_forward.pdf}}
\hspace*{0mm}\fbox{\includegraphics[width=0.49\linewidth]{bpfig_backward.pdf}}
}
\caption{
{\bf Top left.} A multi-layer neural network can distort the
input space to make the classes (whose examples are supposed to
be on the red and blue lines, here) linearly separable. Note how
a regular grid in input space is also transformed. This is an
illustrative example with only two input units, two hidden units
and one output unit, but the networks used for object recognition
or natural language processing contain tens or hundreds of thousand
of units. Reproduced with permission by Chris Olah from~\url{http://colah.github.io/}.
\newline
{\bf Top right.} The chain rule of derivatives tells us how
two small effects (that of a small change of $x$ on $y$, and that of $y$ on $z$)
are composed. A small change $\Delta x$ in $x$
gets transformed first into a small change $\Delta y$ in $y$
by getting multiplied by $\frac{\partial y}{\partial x}$
(that is the definition of partial derivative). Similarly,
the change $\Delta y$ creates a change $\Delta z$ in $z$. Substituting one equation
into the other gives the chain rule of derivatives, i.e., how
$\Delta x$ gets turned into $\Delta z$ through multiplication
by the product of $\frac{\partial y}{\partial x}$ and
$\frac{\partial z}{\partial x}$. It also works when $x$, $y$
and $z$ are vectors (and the derivatives are Jacobian matrices). \newline
{\bf Bottom left.} The equations used for computing the forward pass in a neural net
with two hidden layers. At each layer, we first compute the total input,
$z$, to each unit which is a weighted sum of the outputs of the units in
the layer below. Then a non-linear function $f(.)$ is applied to $z$ to get
the output of the unit.  For simplicity, we have omitted bias terms. \newline
{\bf Bottom right.} The equations used for computing the backward pass. At each hidden
layer we compute the error-derivative w.r.t. the output of each unit which
is a weighted sum of the error-derivatives w.r.t. the total inputs to the
units in the layer above. We then convert the error-derivative w.r.t. the
output into the error-derivative w.r.t. the input by multipling it by the
derivative of $f(z)$.  At the output layer, the error-derivative w.r.t. the
output of a unit is computed by differentiating the cost function. This
gives $y_l-t_l$ If the cost function for unit $l$ is $0.5(y_l-t_l)^2$ where
$t_l$ is the target value.
}
\label{fig:backprop-box}
\end{minipage}
}
\end{figure}

@@

Backpropagation provides an efficient way of computing gradients with respect to all of
the weights in a layered network on a single training case, but there are
still many ways to make use of these gradients. The most obvious method is
to average the gradients over all of the cases in the training set and then
to change each weight in proportion to its negative gradient. This
moves the weight vector in the direction of steepest descent in the
objective function. 
%However, this is needlessly slow when the 
%training set contains millions of cases and is highly redundant
%with most examples conveying no new information about the mapping from
%inputs to outputs.  
However, this is needlessly slow when the 
training set contains millions of cases because each example or small batch of
examples can be seen as a noisy (but correct in average) version of the true total gradient: a very large
number of these noisy steps can converge much faster than very few but expensive clean steps because even
the clean gradient is not the optimal direction in which to go.
%
It is much faster to compute the average gradient
on a small ``mini-batch'' of randomly chosen training cases and to update
each weight by this stochastic estimate of its full gradient, times a
learning rate which decays as learning proceeds. Stochastic gradient
descent can be improved by using the history of the individual weight
gradients to adapt the learning rate for each weight~\citep{bottou-bousquet-2008-small}
or to provide additional increments that depend on earlier gradients 
but these techniques are still the subject of controversy and are beyond
the scope of this review (see~\citet{Montavon2012} for tricks of the trade). 
However, it is now widely accepted that some
form of stochastic gradient descent is the best bet for large, redundant
datasets.

@@

The hidden layers of a multi-layer neural network learn to represent the
network's inputs in a way that makes it easy to predict the target
outputs. This is nicely demonstrated by training a multi-layer neural
network to predict the next word in a sequence from a local context of
earlier words~\citep{BenDucVin01-short}. 
Each context word is presented to the network as a one-of-N
vector in which one component has a value of $1$ and the rest are $0$. The
output layer has one unit per word and is forced to produce a probability
distribution over words by using the ``softmax'' function:
\begin{equation}
p_j = \frac{e^{z_j}}{\sum_i e^{z_i}}
\end{equation}
where $z_j$ is the total input received by output unit $j$. The hidden
layers use a non-linearity such as the logistic, tanh, or rectifier, respectively:
\begin{equation}
y_j = \frac{1}{1+ e^{z_j}}, \ \ \ y_j = \frac{e^{z_j}-e^{-z_j}}{e^{z_j}+e^{-z_j}}, \ \ \
 y_j = \max(0, z_j)
\end{equation}

The connectivity of the network is restricted so that each of the one-of-N
input vectors must first be converted to a ``word-vector'' before being
used by subsequent layers to give a high probability to the target next
word.  The network learns word-vectors that contain many active components
each of which can be interpreted as a separate feature of the word. This
was first demonstrated using sequences of three words, such as
(Colin, Mother, Victoria), derived from two family trees~\citep{RHW}. The learned word
vector for Colin contained ``semantic'' features representing which family
tree Colin was in, what branch of that tree he was in, and what generation
he was from. 
%% TWO NATIVE ENGLISH SPEAKERS AGREE THAT THE "FROM" SHOULD STAY AT THE
%% END IN ORDER NOT TO SOUND STILTED.
These semantic features were not explicitly present
in the input.  They were discovered by the learning procedure as a good way
of factorizing the structured relationships between the input and output
symbols into multiple ``micro-rules''.  For example, given Colin's
generation and the learned feature of Mother that the output should be one
generation higher than the input, we can already restrict the answer to
people in a specific generation. The ability of deep learning to discover
many such micro-rules allows it to generalize well to novel combinations of
individually familiar context words.

The family tree demonstration used a small, clean dataset in which the
regularities could be captured by micro-rules that were never violated and
the features were easy to interpret, but learned word-vectors also work
very well when the word sequences come from a large corpus of real text and
the individual micro-rules are unreliable~\citep{BenDucVin01-short}. When trained to
predict the next word in a news story, for example, the learned word
vectors for Monday and Tuesday are very similar, as are the word vectors
for Vietnam and Iraq.  Vector representations of words learned from text
are now very widely used in natural language 
applications~\citep{Schwenk-2007,collobert:2011b,Socher-2011,Mikolov-et-al-NIPS2013}.

@@

The architecture of a typical ConvNet is 
%shown in
%figure~\ref{fig-convnet}. 
structured as a series of stages. The first few stages are composed of two
types of layers: convolutional layers and pooling layers. Units in a
convolutional layer are organized in {\em feature maps}, within which
each unit is connected to local patches in the feature maps of the
previous layer through a set of weights called a {\em filter
  bank}. The result of this local weight sum is then passed through a
non-linearity.  All units in a feature map share the same filter
bank. Different feature maps in a layer use different filter banks. The
reason for this architecture is twofold. First, in array data such as
images, local groups of values are often highly correlated forming
local motifs that are easily detected. Second, the local statistics of
images and other signals are invariant to location. In other word, if
a motif can appear in one part of the image, it could appear anywhere,
hence the idea of units at different locations sharing the same
weights.  Mathematically, the filtering operation performed by a
feature map is a {\em discrete convolution}, hence the name
ConvNet.

@@

This architecture exploits the property that many natural signals are
{\em compositional hierarchies}. In images, local combinations of
edges form motifs, motifs assemble into parts, and parts form objects.
Similar hierarchies exists in speech and text from sounds to phones,
phonemes, syllables, words, and sentences. The features that form
elements at the next level can vary in relative position and
appearance. The convolutional and pooling layers in ConvNets are
directly inspired by the classic notions of simple cells and complex
cells in visual neuroscience~\cite{Hubel62}, and the overall
architecture is reminiscent of visual cortex ventral pathway hierarchy
LGN-V1-V2-V4-IT~\citep{Felleman+VanEssen-1991}. ConvNets have their roots in
Fukushima's Neocognitron~\citep{fukushima-82} whose architecture was
somewhat similar but did not have an end-to-end supervised learning
algorithm such as backpropagation+SGD.

There have been numerous applications of convolutional networks going
back to the early 1990s. A major early success
%% THIS REWORDING AVOIDS COMPLAINTS FROM HECHT NIELSEN ETC.
was a bank check
reading system which was commercially deployed by AT\&T and NCR in
1994~\cite{lecun-98}. The system used a ConvNets trained jointly with
a graphical model that implemented language constraints. By the late
1990s this system was reading over 10\% of all the checks in the
US. Microsoft later deploy ConvNets in a number of OCR and handwriting
recognition
systems~\cite{simard-03,chellapilla-ist-06,chellapilla-iwfhr-06b}
including for Arabic and
Chinese~\cite{abdulkader-iwfhr-06,chellapilla-iwfhr-06a}. ConvNets
have also been used since the early 1990s for object detection in
natural images, including faces and
pedestrians~\cite{vaillant-monrocq-lecun-94,garcia-delakis-04,osadchy-07,nasse-09,sermanet-cvpr-13}.

When backpropagation was first introduced, its most exciting use was for
training recurrent neural networks. These receive input vectors at each
time-step and have one or more layers of hidden units that, in addition to
having connections to higher layers or ouput units, also have connections
that allow the neurons within a layer to receive weighted input from those
same neurons at the previous time-step.  As illustrated in
Figure~\ref{fig:rnn}, the gradient with respect to a recurrent
weight, $w_{ij}$ that allows neuron $i$ at time $t$ to influence the
activity of neuron $j$ at time $t+1$ can be computed by
backpropagation-through-time (BPTT) of the error derivative for $j$ at time
$t+1$. The full gradient with respect to $w_{ij}$ on a whole input sequence is just the
sum of its gradients at each time-step.

In the 1990s, the machine learning community decided that backpropagation
in deep neural networks did not work well in practice and it was largely
abandoned.  At the time, people believed that learning many layers of
non-linear feature detectors with no prior knowledge was much too difficult
for a simple gradient descent procedure and that the networks were getting
trapped in poor local minima. We now know that the main reasons for the
limited success of backpropagation in deep networks were quite different:
The labeled datasets were far too small, the computers were far too slow,
and the weights were not initialised sensibly~\citep{GlorotAISTATS2010-small}.

The obvious weakness of gradient descent is that it can get trapped in poor
local minima, but in high-dimensional weight-spaces this is much less
common than getting stuck close to saddle points where the error surface
curves upwards in most directions but downwards in a few. Recent
results~\citep{Dauphin-et-al-NIPS2014-small,Choromanska-et-al-AISTATS2015} strongly suggest
the vast majority of the critical points (where derivatives of the total
objective function with respect to the parameters vanish) are saddle points, not
local minima, and that in large, deep networks almost all of the 
saddle points with predominantly upwards curvature  
give very similar performance so it does not matter which one we find.  This goes a long way
towards explaining why stochastic gradient descent works so well for deep
neural networks.

Interest in deep feed-forward networks was revived in
%% Even though I had a conference  paper in 2005, the revival was in 2006!
2006~\citep{IJCAI,Hinton06-small,Bengio-nips-2006-small,ranzato-07-small} 
by the introduction
of unsupervised learning procedures that could create layers of feature
detectors without requiring labeled data. The objective in learning each
layer of feature detectors was to be able to reconstruct or model the
activities of feature detectors (or raw inputs) in the layer below.  By
``pre-training'' several layers of progressively more complex feature
detectors using this reconstruction objective, the weights of a deep
network could be initialized to sensible values.  A final layer of output
units could then be added to the top of the network and the whole deep
system could be fine-tuned using standard
backpropagation \citep{Hinton-Science2006}. This worked remarkably well for recognizing
handwritten digits especially when the amount of labeled data was very
limited~\citep{Bengio-nips-2006-small,Erhan+al-2010-small}.

The first major application of this pre-training approach was in speech
recognition and it was made possible by the advent of fast Graphics
Processing Units (GPUs) that were convenient to
program~\citep{RainaICML09-small}.  In 2009, the approach was used to map
short temporal windows of coefficients extracted from a soundwave to a set
of probabilities for the various fragments of speech that might be
represented by the frame in the center of the window.  It achieved
record-breaking results on a standard speech recognition benchmark that
used a small vocabulary~\citep{TIMITpaper} and was quickly developed to
give record-breaking results on a large vocabulary task~\citep{Dahl2012}.
By 2012, versions of the deep net from 2009 were being developed by many of
the major speech groups~\citep{Hinton-et-al-2012} and were already being
deployed in Android phones.  For smaller datasets, unsupervised
pre-training helps to prevent overfitting~\citep{Erhan+al-2010-small},
leading to significantly better generalization when the number of labeled
examples is small, or in a transfer setting where we have lots of
examples for some ``source'' tasks but very few for some ``target''
task. In this context, unsupervised representation learning helped to win
several transfer learning
competitions~\citep{UTLC+LISA-2011-small,Goodfellow-icml2012}.  Once deep
learning had been rehabilitated, it turned out that large labeled datasets
did not need the pre-training stage, provided the scales of the initial
random weights were carefully chosen to avoid very large or very small
initial gradients~\citep{GlorotAISTATS2010-small} and that even better
results (especially for very deep networks)
could be obtained by using rectifying non-linearities rather than
sigmoid or tanh non-linearities
~\citep{Nair-2010-small,Glorot+al-AI-2011-small,Krizhevsky-2012-small,Dahl-et-al-ICASSP2013}.

Interest in recurrent networks (which are deep in time) was revived by
``Long Short Term Memory'' networks (LSTMs) that use special hidden units
whose natural behaviour is to remember inputs for a long
time~\citep{Hochreiter+Schmidhuber-1997}.  
A linear ``memory cell'' interacts with other units
via gated connections whose weights are multiplied by the state of a
logistic gating unit that has an activation between $0$ and $1$. The input
vector has little effect on the state of that cell if the cell's input
gate has a low activation so memory cells can learn to selectively ignore
the input.  Similarly, a memory cell has little effect on the output if its
output gate has low activation. Each memory cell is a gated leaky neuron: 
it has a connection to
itself at the next time-step that has a weight of 1 so it copies its
own real-valued state and accumulates the external signal, 
but this self-connection is {\em gated} by a ``forget gate'' so
the stored value quickly leaks away unless the forget gate has an activation
near $1$.  The gates themselves have their activities determined by
weighted inputs that they receive from the input vectors and all of the
memory cells.

When LSTMs were first described % in English
\citep{Hochreiter+Schmidhuber-1997},
their complexity may have been off-putting
and, despite their impressive ability to learn very long-range temporal
structure on toy problems, they were largely ignored until they achieved
excellent results in a competition to read cursive
hand-writing~\citep{Graves-et-al-2009}. They have subsequently proved to be very
effective for tasks that have sequential input.  When they use several
hidden layers, they are better than
deep feed-forward networks for speech recognition~\citep{Graves-et-al-ICASSP2013} and
unlike feed-forward nets they can implement an entire speech recognition
system that goes all the way from coefficents derived from the soundwave to
the sequence of characters in the transcription.  LSTMs or related forms of gated
units are also what is currently used for the encoder and decoder networks that
perform so well at machine
translation~\citep{Bahdanau-et-al-arxiv2014,Sutskever-et-al-NIPS2014}.

@@

>>>>

So far, most applications of deep learning use feedforward neural
networks, illustrated in Figure~\ref{fig:backprop-box}, which learn
to map a fixed-size input (e.g., an image) to a fixed-size output
(e.g. a probability for each one of several candidate object categories). 
They do so by successive transformations corresponding to different
layers of artificial neurons (called {\em hidden units}),
which can be interpreted as learned derived features of the raw input. However, 
for tasks that involve sequential inputs, such as speech
and language, it is often better to use recurrent neural
networks, illustrated in Figure~\ref{fig:rnn} and discussed in Section~\ref{sec:rnn}. 
Recurrent neural networks process an input sequence one element at a time,
maintaining in their hidden units a ``state vector'' that implicitly contains information
about the history all the past elements of the sequence.
Recurrent networks are very good at predicting the next
character in text~\citep{Sutskever-et-al-ICML2011} or the next
word in a sequence~\citep{Mikolov-et-al-NIPS2013} but they can also be
used for more complex tasks.  After reading an English sentence one
word at a time, for example, an English ``encoder'' network can be
trained so that the final state vector of its hidden units is a good
representation of the thought expressed by the sentence.  This thought
vector can then be used as the initial hidden state of (or as extra
input to)  a jointly trained French ``decoder'' network which outputs a
probability distribution for the first word of the French
translation. If a particular first word is chosen 
from this distribution and provided as input to the decoder network it will
then output a probability distribution for the second word of the
translation and so on until a full stop is
chosen~\citep{Bahdanau-et-al-arxiv2014,Sutskever-et-al-NIPS2014}.
Overall, this process generates sequences of French words according to
a probability distribution that depends on the English sentence.

@@

\begin{figure}[htp]
\fbox{
\begin{minipage}{\textwidth}
\centerline{
\fbox{\includegraphics[width=0.69\linewidth]{netvis-simple-2S.png}}
\fbox{\includegraphics[width=0.28\linewidth]{chain-rule.pdf}}}
\centerline{
\hspace*{.5mm}\fbox{\includegraphics[width=0.49\linewidth]{bpfig_forward.pdf}}
\hspace*{0mm}\fbox{\includegraphics[width=0.49\linewidth]{bpfig_backward.pdf}}
}
\caption{
{\bf Top left.} A multi-layer neural network can distort the
input space to make the classes (whose examples are supposed to
be on the red and blue lines, here) linearly separable. Note how
a regular grid in input space is also transformed. This is an
illustrative example with only two input units, two hidden units
and one output unit, but the networks used for object recognition
or natural language processing contain tens or hundreds of thousand
of units. Reproduced with permission by Chris Olah from~\url{http://colah.github.io/}.
\newline
{\bf Top right.} The chain rule of derivatives tells us how
two small effects (that of a small change of $x$ on $y$, and that of $y$ on $z$)
are composed. A small change $\Delta x$ in $x$
gets transformed first into a small change $\Delta y$ in $y$
by getting multiplied by $\frac{\partial y}{\partial x}$
(that is the definition of partial derivative). Similarly,
the change $\Delta y$ creates a change $\Delta z$ in $z$. Substituting one equation
into the other gives the chain rule of derivatives, i.e., how
$\Delta x$ gets turned into $\Delta z$ through multiplication
by the product of $\frac{\partial y}{\partial x}$ and
$\frac{\partial z}{\partial x}$. It also works when $x$, $y$
and $z$ are vectors (and the derivatives are Jacobian matrices). \newline
{\bf Bottom left.} The equations used for computing the forward pass in a neural net
with two hidden layers (each composed of a group of hidden units that are
similarly connected). At each layer, we first compute the total input,
$z$, to each unit which is a weighted sum of the outputs of the units in
the layer below. Then a non-linear function $f(.)$ is applied to $z$ to get
the output of the unit.  For simplicity, we have omitted bias terms. \newline
{\bf Bottom right.} The equations used for computing the backward pass. At each hidden
layer we compute the error-derivative w.r.t. the output of each unit which
is a weighted sum of the error-derivatives w.r.t. the total inputs to the
units in the layer above. We then convert the error-derivative w.r.t. the
output into the error-derivative w.r.t. the input by multipling it by the
derivative of $f(z)$.  At the output layer, the error-derivative w.r.t. the
output of a unit is computed by differentiating the cost function. This
gives $y_l-t_l$ If the cost function for unit $l$ is $0.5(y_l-t_l)^2$ where
$t_l$ is the target value.
}
\label{fig:backprop-box}
\end{minipage}
}
\end{figure}

@@

Backpropagation provides an efficient way of computing gradients with respect to all of
the weights in a layered network on a single training case, but there are
still many ways to make use of these gradients. The most obvious method is
to average the gradients over all of the cases in the training set and then
to change each weight in proportion to its negative gradient. This
moves the weight vector in the direction of steepest descent in the
objective function. 
%However, this is needlessly slow when the 
%training set contains millions of cases and is highly redundant
%with most examples conveying no new information about the mapping from
%inputs to outputs.  
However, this is needlessly slow when the 
training set contains millions of cases because each example or small batch of
examples can be seen as a noisy (but correct in average) version of the true total gradient: a very large
number of these noisy steps can converge to a good solution
much faster than very few but expensive clean steps, because even
the clean gradient is not the optimal direction in which to go.
%
It is much faster to proceed one small ``mini-batch'' examples at a time,
i.e., compute the average gradient
on a small set of randomly chosen training cases (say 128) and to update
each weight by this stochastic estimate of its full gradient, times a
learning rate which decays as learning proceeds. Stochastic gradient
descent can be improved by using the history of the individual weight
gradients to adapt the learning rate for each weight~\citep{bottou-bousquet-2008-small}
or to provide additional increments that depend on earlier gradients 
but these techniques are still the subject of controversy and are beyond
the scope of this review (see~\citet{Montavon2012} for tricks of the trade). 
However, it is now widely accepted that some
form of stochastic gradient descent is the best bet for large
datasets.

@@

The hidden layers of a multi-layer neural network learn to represent the
network's inputs in a way that makes it easy to predict the target
outputs. This is nicely demonstrated by training a multi-layer neural
network to predict the next word in a sequence from a local context of
earlier words~\citep{BenDucVin01-short}. 
Each context word is presented to the network as a one-of-N
vector, i.e., in which one component has a value of $1$ and the rest are $0$. The
output layer has one unit per word and is forced to produce a probability
distribution over words by using the ``softmax'' function:
\begin{equation}
p_j = \frac{e^{z_j}}{\sum_i e^{z_i}}
\end{equation}
where $z_j$ is the total input received by output unit $j$. 
\iffalse
The hidden
layers use a non-linearity such as the logistic, tanh, or rectifier, respectively:
\begin{equation}
y_j = \frac{1}{1+ e^{z_j}}, \ \ \ y_j = \frac{e^{z_j}-e^{-z_j}}{e^{z_j}+e^{-z_j}}, \ \ \
 y_j = \max(0, z_j).
\end{equation}
\fi

The presence of a particular word at the $i$-th position in the input of the network 
sets the value of specialized hidden units to values that correspond to the
representation of that word. Different words create a different pattern of
activations, or word-vector, illustrated in Figure~\ref{fig:word-embeddings}.
In a language model, the other layers of the network learn to convert the input word-vectors
into an output word-vector for the predicted next-word, which can be used
to predict the probability for any word in the vocabulary to appear as the next word.
The network learns word-vectors that contain many active components
each of which can be interpreted as a separate feature of the word. This
was first demonstrated using sequences of three words, such as
(Colin, Mother, Victoria), derived from two family trees~\citep{RHW}. The learned word
vector for Colin contained ``semantic'' features representing which family
tree Colin was in, what branch of that tree he was in, and what generation
he was from. 
%% TWO NATIVE ENGLISH SPEAKERS AGREE THAT THE "FROM" SHOULD STAY AT THE
%% END IN ORDER NOT TO SOUND STILTED.
These semantic features were not explicitly present
in the input.  They were discovered by the learning procedure as a good way
of factorizing the structured relationships between the input and output
symbols into multiple ``micro-rules''.  For example, given Colin's
generation and the learned feature of Mother that the output should be one
generation higher than the input, we can already restrict the answer to
people in a specific generation. The ability of deep learning to discover
many such micro-rules allows it to generalize well to novel combinations of
individually familiar context words.

The family tree demonstration used a small, clean dataset in which the
regularities could be captured by micro-rules that were never violated and
the features were easy to interpret, but learned word-vectors also work
very well when the word sequences come from a large corpus of real text and
the individual micro-rules are unreliable~\citep{BenDucVin01-short}. When trained to
predict the next word in a news story, for example, the learned word
vectors for Tuesday and Wednesday are very similar, as are the word vectors
for Vietnam and Iraq. This is not to say that Tuesday and Wednesday are
put in the same category, but rather that they share many semantic and
grammatical attributes, so that what the neural network learns from one day of
the week can often be generalized to the other, while individual days remain
different (for example one imagines that the neural network also
learns to some features that distinguises week-end days and weekdays, while
Saturday and Sunday might have special roles in some religions, etc.).
Such representations are called {\em distributed representations} because
its elements (the features) are not mutually exclusive and there exists
a rich and vast set of possible configurations of values of these features,
corresponding to the variations seen in the observed data.
These word-vectors are thus {\em learned features} that were not determined
ahead of time by humans, but automatically discovered by the neural network.
Vector representations of words learned from text
are now very widely used in natural language 
applications~\citep{Schwenk-2007,collobert:2011b,Socher-2011,
Mikolov-et-al-NIPS2013,Bahdanau-et-al-arxiv2014,Sutskever-et-al-NIPS2014}.

@@

The architecture of a typical ConvNet is 
%shown in
%figure~\ref{fig-convnet}. 
structured as a series of stages. The first few stages are composed of two
types of layers: convolutional layers and pooling layers. Units in a
convolutional layer are organized in {\em feature maps}, within which
each unit is connected to local patches in the feature maps of the
previous layer through a set of weights called a {\em filter
  bank}. The result of this local weight sum is then passed through a
non-linearity.  All units in a feature map share the same filter
bank. Different feature maps in a layer use different filter banks. The
reason for this architecture is twofold. First, in array data such as
images, local groups of values are often highly correlated forming
local motifs that are easily detected. Second, the spatially local statistics of
images and other signals are invariant to location. In other word, if
a motif can appear in one part of the image, it could appear anywhere,
hence the idea of units at different locations sharing the same
weights.  Mathematically, the filtering operation performed by a
feature map is a {\em discrete convolution}, hence the name
ConvNet.

@@

This architecture exploits the property that many natural signals are
{\em compositional hierarchies}. In images, local combinations of
edges form motifs, motifs assemble into parts, and parts form objects.
Similar hierarchies exists in speech and text from sounds to phones,
phonemes, syllables, words, and sentences. The features that form
elements at the next level can vary in relative position and
appearance. The convolutional and pooling layers in ConvNets are
directly inspired by the classic notions of simple cells and complex
cells in visual neuroscience~\citep{Hubel62}, and the overall
architecture is reminiscent of the LGN-V1-V2-V4-IT visual cortex ventral pathway hierarchy
~\citep{Felleman+VanEssen-1991}. ConvNets have their roots in
Fukushima's Neocognitron~\citep{fukushima-82} whose architecture was
somewhat similar but did not have an end-to-end supervised learning
algorithm such as backpropagation plus stochastic gradient descent.

There have been numerous applications of convolutional networks going
back to the early 1990s. A major early success
%% THIS REWORDING AVOIDS COMPLAINTS FROM HECHT NIELSEN ETC.
was a bank check
reading system which was commercially deployed by AT\&T and NCR in
1994~\citep{lecun-98}. The system used a ConvNet trained jointly with
a probabilistic model that implemented language constraints. By the late
1990s this system was reading over 10\% of all the checks in the
US. Microsoft later deployed ConvNets in a number of OCR and handwriting
recognition
systems~\citep{simard-03,chellapilla-ist-06,chellapilla-iwfhr-06b}
including for Arabic and
Chinese~\citep{abdulkader-iwfhr-06,chellapilla-iwfhr-06a}. ConvNets
have also been used since the early 1990s for object detection in
natural images, including faces and
pedestrians~\cite{vaillant-monrocq-lecun-94,garcia-delakis-04,osadchy-07,nasse-09,sermanet-cvpr-13}.

@@

When backpropagation was first introduced, its most exciting use was for
training recurrent neural networks. These receive input vectors at each
time-step and have one or more layers of hidden units that, in addition to
having connections to higher layers or ouput units, also have connections
that allow the neurons within a layer to receive weighted input from those
same neurons at the previous time-step.  As illustrated in
Figure~\ref{fig:rnn}, when we consider the outputs of the hidden units at
different discrete time steps as if they were the outputs of different neurons
in a deep multi-layer network, it becomes clear how we can apply backpropagation
to train recurrent networks. The
the gradient with respect to a recurrent
weight, $w_{ij}$ that allows neuron $i$ at time $t$ to influence the
activity of neuron $j$ at time $t+1$ can be computed by
backpropagation-through-time (BPTT) of the error derivative for $j$ at time
$t+1$. The full gradient with respect to $w_{ij}$ on a whole input sequence is just the
sum of its gradients at each time-step.

In the 1990s, the machine learning community decided that backpropagation
in deep neural networks did not work well in practice and it was largely
abandoned.  At the time, people believed that learning many layers of
non-linear feature detectors with no prior knowledge was much too difficult
for a simple gradient descent procedure and that the networks were getting
trapped in poor local minima, weight configurations for which no small
change would reduce the average error. We now know that the main reasons for the
limited success of backpropagation in deep networks were quite different:
The labeled datasets were far too small, the computers were far too slow,
and the weights were not initialised sensibly~\citep{GlorotAISTATS2010-small}.

The obvious perceived weakness of gradient descent is that it can get trapped in poor
local minima, but in high-dimensional weight-spaces this is much less
common than getting stuck close to {\em saddle points}, where the error surface
curves upwards in most directions but downwards in a few. Recent
results~\citep{Dauphin-et-al-NIPS2014-small,Choromanska-et-al-AISTATS2015} strongly suggest
that the vast majority of the critical points\footnote{where derivatives of the total
objective function with respect to the parameters vanish, and include minima,
maxima and saddle points} are saddle points, not
local minima, and that in large and deep networks almost all of the 
saddle points with predominantly upwards curvature  
give very similar performance, so it does not matter which one we find.  This goes a long way
towards explaining why stochastic gradient descent works so well for deep
neural networks.

Interest in deep feedforward networks was revived in
%% Even though I had a conference  paper in 2005, the revival was in 2006!
2006~\citep{IJCAI,Hinton06-small,Bengio-nips-2006-small,ranzato-07-small} 
by the introduction
of unsupervised learning procedures that could create layers of feature
detectors without requiring labeled data. The objective in learning each
layer of feature detectors was to be able to reconstruct or model the
activities of feature detectors (or raw inputs) in the layer below.  By
``pre-training'' several layers of progressively more complex feature
detectors using this reconstruction objective, the weights of a deep
network could be initialized to sensible values.  A final layer of output
units could then be added to the top of the network and the whole deep
system could be fine-tuned using standard
backpropagation \citep{Hinton-Science2006,Bengio-nips-2006-small,ranzato-07-small}. 
This worked remarkably well for recognizing
handwritten digits especially when the amount of labeled data was very
limited~\citep{Erhan+al-2010-small}.

The first major application of this pre-training approach was in speech
recognition and it was made possible by the advent of fast Graphics
Processing Units (GPUs) that were convenient to
program~\citep{RainaICML09-small} and allowed researchers to train
networks 10 or 20 times faster.  In 2009, the approach was used to map
short temporal windows of coefficients extracted from a soundwave to a set
of probabilities for the various fragments of speech that might be
represented by the frame in the center of the window.  It achieved
record-breaking results on a standard speech recognition benchmark that
used a small vocabulary~\citep{TIMITpaper} and was quickly developed to
give record-breaking results on a large vocabulary task~\citep{Dahl2012}.
By 2012, versions of the deep net from 2009 were being developed by many of
the major speech groups~\citep{Hinton-et-al-2012} and were already being
deployed in Android phones.  For smaller datasets, unsupervised
pre-training helps to prevent overfitting~\citep{Erhan+al-2010-small},
leading to significantly better generalization when the number of labeled
examples is small, or in a transfer setting where we have lots of
examples for some ``source'' tasks but very few for some ``target''
task. In this context, unsupervised representation learning helped to win
several transfer learning
competitions~\citep{UTLC+LISA-2011-small,Goodfellow-icml2012}.  Once deep
learning had been rehabilitated, it turned out that large labeled datasets
did not need the pre-training stage, provided the scales of the initial
random weights were carefully chosen to avoid very large or very small
initial gradients~\citep{GlorotAISTATS2010-small}. Even better
results, especially for very deep networks,
could be obtained by using rectifying non-linearities rather than
sigmoid or tanh non-linearities
~\citep{Nair-2010-small,Glorot+al-AI-2011-small,Krizhevsky-2012-small,Dahl-et-al-ICASSP2013}.

Interest in recurrent networks (which correspond to a very deep network
when unfolded in time as in Figure~\ref{fig:rnn}) was revived by
``Long Short Term Memory'' networks (LSTMs) that use special hidden units
whose natural behaviour is to remember inputs for a long
time~\citep{Hochreiter+Schmidhuber-1997}.  
A linear ``memory cell'' interacts with other units
via gated connections whose weights are multiplied by the state of a
logistic gating unit that takes a value between $0$ and $1$. An input gating
unit can learn when to isolate the cell from the effect of the outside world,
and an output gating unit prevents the content of the memory cell to be
broadcast to the rest of the neural network. More importantly, the
memory cell itself acts like an accumulator or a gated leaky neuron:  it has a connection to
itself at the next time-step that has a weight of 1 so it copies its
own real-valued state and accumulates the external signal, 
but this self-connection is {\em gated} by a ``forget gate'' so
the stored value quickly leaks away unless the forget gate has an activation
near $1$.  The gates themselves have their activities determined by
weighted inputs that they receive from the input vectors and all of the
memory cells.

When LSTMs were first described % in English
\citep{Hochreiter+Schmidhuber-1997},
their complexity may have been off-putting
and, despite their impressive ability to learn very long-range temporal
structure on toy problems, they were largely ignored until they achieved
excellent results in a competition to read cursive
hand-writing~\citep{Graves-et-al-2009}. They have subsequently proved to be very
effective for tasks that have sequential input.  When they use several
hidden layers, they are better than
deep feedforward networks for speech recognition~\citep{Graves-et-al-ICASSP2013} and
unlike feedforward nets they can implement an entire speech recognition
system that goes all the way from coefficents derived from the soundwave to
the sequence of characters in the transcription.  LSTMs or related forms of gated
units are also what is currently used for the encoder and decoder networks that
perform so well at machine
translation~\citep{Bahdanau-et-al-arxiv2014,Sutskever-et-al-NIPS2014}.
